---
title: "Final Report"
author: "Group 5: Megan Lui, Michelle Wong, Yael Yossefy, Dirgh Shah"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Introduction

### Scientific Question:

Does the relationship between EtCO2 and AMC behave differently in the
presence of relevant medical history?

### Clinical relevance:

Observing the relationship between EtCO2 (end-tidal carbon dioxide) and
AMC (area of maximum compression) is crucial to understanding the
quality of chest compression during CPR (cardiopulmonary resuscitation),
a life-saving procedure during cardiac arrests. Chest compressions are
crucial to reviving a patient or achieving ROSC or Return to Spontaneous
Circulation, though they requires adequate rate, depth, and location for
the best results. However, CPR guidelines are not as precise, as
differences in body size and histories of chronic disease can alter the
heart's location. When patients have compression in the aortic root,
there is an obstruction of blood flow from the heart and can limit
effective perfusion and worsen the prognosis. Compressions over the left
ventricle or LV can improve the blood circulation and lead to
higher-quality CPR. This relationship between EtCO2 (where higher levels
have been shown to correlate with better blood flow from past studies)
and AMC may differ dependent on a patient's history, as these conditions
can impact the ventilation-perfusion mismatch or other presenting
symptoms.

## Methods

### Design of Study:

The metabolic syndrome and cognitive function study is a retrospective
observational analysis which focuses on investigating the relationship
between EtCO2 and the AMC during CPR and whether this relationship is
affected by the presence of relevant medical histories of patients.

This study focuses on 2 primary areas of compression, the left ventricle
(LV) and the left ventricle outflow tract (LVOT)/aortic root, and how
compression over these AMCs impact EtCO2 levels and whether this impact
changes by the CPR patients' pre existing medical conditions, such as
coronary artery disease, congestive hearth failure, chronic kidney
disease, diabetes mellitus and myocardial infarction.

### Data set Details:

The data set consists of 129 cases of in-hospital cardiac arrests where
transesophageal echocardiography (TEE) imaging was employed during CPR.
This advanced imaging technique is used to identify the specific area of
the chest targeted for compression to precisely classify the AMC - LV or
LVOT.

Key variables in the data set include:

-   Dependent Variable:

    -   EtCO2: Measured during CPR, represents levels of exhaled CO2 and
        serves as a proxy for cardiac output.

-   Independent Variables:

    -   AMC: Areas of compression, classified as either LV or LVOT.

    -   Medical History: Binary indicators for conditions like coronary
        artery diseas (hist1), congestive hearth failure (hist2),
        chronic kidney disease (hist3), diabetes mellitus (hist4), and
        myocardial infarction (hist5).

-   Covariates:

    -   Demographics like race and gender.

    -   TEE operator experience level (e.g. attending physicain,
        resident, etc.).

    -   Type of CPR performed (manual, mechanical, alternating).

    -   Timing of cardiac arrest (known or approximated).

First, the data set underwent pre-processing and cleaning, which
included the removal of invalid data entries. The next step involved
converting the numeric values flagged as outliers (e.g. placeholder
values like 9999) to missing (NA) values and excluding them from the
analysis.

### Cohort:

The study cohort consisted of adult patients who had experienced
in-hospital cardiac arrest and received CPR where TEE was used. The
inclusion criteria required proper TEE imaging data to identify the
precise location of AMC and valid measurements of EtCO2 levels. patients
with any significant missing data were excluded. After pre-processing
and cleaning, the data set included 106 valid observations (patients)
for further analysis.

### Analysis Methodology:

A multivariable linear model was implemented to analyse the relationship
between EtCO2 level and AMC. Interaction terms between AMC the the
medical history variables were used in this model to assess whether the
relationship between EtCO2 and AMC varied by the presence of 5 specific
pre-existing medical conditions.

The analysis included two models:

-   A reduced model, which included only the main effects of AMC and
    medical history.

-   A full model, which included interaction terms to evaluate whether
    medical history affected the relationship between EtCO2 and AMC.

An Analysis of Variance (ANOVA) test was used to compare these models
and evaluate the contribution of the interaction terms to model fit.

### Rationale:

A linear regression model was implemented based on the assumption that
the relationship between EtCO2 and AMC is linear, as CO2 levels in the
blood reflect cardiac output which is influnced by the area of
compression (LV or LVOT). We assume higher EtCO2 levels when the AMC is
LV. This specific approach allows for testing the main effects while
including interaction terms to study more complex relationships. The
inclusion of these interaction terms is very crucial for identifying
whether specif medical histories alter the impact of AMC on EtCO2.

Assumptions include:

-   Linearity and Homoscedasticity: Checked via residual plots.

-   Independence: Verified using the Durbin-Watson test.

-   Normality of Residuals: Assessed through Q-Q plots.

-   Multicollinearity: Evaluated using Variance Inflation Factors (VIF),
    with all predictors having suitable values (\< 5).

This methodological approach addresses potential confounding factors and
statistical validity while guaranteeing a thorough investigation of the
research question.

```{r}
library(readxl)
library(mosaic)
data <- read_excel("cardiac_arrests.xlsx")
#head(data)
```

### Cleaning

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

data <- data %>%
  mutate(across(where(is.numeric), ~ replace(.x, .x == 9999, NA))) 

data$race <- factor(data$race, levels = c(1, 2, 3, 4, 99), 
                    labels = c("White", "Black/African American", 
                               "Asian", "Latinx", "Other"))
data$amc <- factor(data$amc, levels = c(1, 2), 
                   labels = c("LV", "LVOT/Aortic root"))
data$op_level <- factor(data$op_level, levels = c(1, 2, 3, 99), 
                        labels = c("Attending physician", "Fellow", 
                                   "Resident", "Other"))
data$cpr_type <- factor(data$cpr_type, levels = c(1, 2, 3), 
                        labels = c("Manual", "Mechanical", "Alternating"))
data$hist1 <- factor(data$hist1, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist2 <- factor(data$hist2, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist3 <- factor(data$hist3, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist4 <- factor(data$hist4, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist5 <- factor(data$hist5, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$rosc <- factor(data$rosc, levels = c(0, 1), 
                    labels = c("No", "Yes"))

data$survival <- factor(data$survival, levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$time_approx <- factor(data$time_approx, levels = c(0, 1), 
                           labels = c("Known", "Approximated"))

#summary(data)
cleaned_data <- na.omit(data)
```

### Assumptions Made:
1.  Linearity and Homoscedasticity (Constant Variance) The relationship
    between the predictors and the outcome are linear. This means that changes in
    the predictors produce proportional changes in the
    outcome.For homoscedasticity, the variance of the residuals
    should be constant across all levels of the independent variables.

```{r}
normal_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist2 + amc:hist3 + 
                          amc:hist4 + amc:hist5, 
                        data = cleaned_data)

plot(normal_model, which = 1)
```

2.  Independence: The residuals from the model appear to be
    independent. This means that the error associated with one
    observation is not correlated with the error of another observation.
    A value close to 2 indicates no autocorrelation, and we get a DW value of 1.9764.

```{r}
library(lmtest)
dwtest(normal_model)
```

3.  Normality of Residuals: The residuals should be normally
    distributed, and we see from the qqplot that they follow the line of best fit. 

```{r}
plot(normal_model, which = 2)
```

4.  No Perfect Multicollinearity: The independent variables should not
    be highly correlated with each other. If they are, it becomes
    difficult to isolate the individual effects of each predictor. In
    ANOVA, this is typically not an issue if the independent variables
    are categorical and factors, but in multiple regression, it could be.

VIF \< 5: Acceptable (low multicollinearity). 

VIF between 5 and 10: Moderate multicollinearity, may warrant investigation. 

VIF \> 10:Problematic (high multicollinearity), often considered a "bad" value.

We see here in our results that we do not have any values greater than 5, which indicates to us that our variables are independent with low multicollinearity. 

```{r}
library(car)
vif(normal_model, 'predictor')
vif(normal_model)
```

### Early linear Regression Model with all possible interaction terms:

Since the assumptions hold, we are able to use a linear regression model to examine the relationships between our variables. 
```{r}
summary(normal_model)
```

## Results

### Tables Figures Statistics

First, let's look at the relationship between AMC and EtCO2, without accounting for different histories of disease.

```{r}
ggplot(cleaned_data, aes(x = amc, y = etco2))+
  geom_boxplot() + 
  labs(x = "AMC", y = "EtCO2", title = "BoxPlots of EtCO2 vs. AMC") + 
  ylim(0,30)
```

```{r}
library(mosaic)
favstats(cleaned_data$etco2[cleaned_data$amc=="LV"])
favstats(cleaned_data$etco2[cleaned_data$amc=="LVOT/Aortic root"])

#library(dplyr)
#library(mosaic)

#cleaned_data |>
#  group_by(amc) |>
#  favstats(etco2)
```

```{r}
ggplot(cleaned_data, aes(x = etco2, fill = amc)) +
  geom_density(alpha = 0.5) + 
  labs(x = "EtCO2", y = "Density", fill = "AMC")

```

Now let's look at how history of disease affects this relationships:

```{r}
library(gridExtra)

plots <- lapply(c("hist1", "hist2", "hist3", "hist4", "hist5"), function(hist_var) {
  ggplot(cleaned_data, aes(x = amc, y = etco2, fill = get(hist_var))) +
  geom_boxplot(outlier.shape = NA) +
  labs(x = "AMC", y = "EtCO2", fill = paste(hist_var)) +
  scale_fill_manual(values = c("lightblue", "salmon")) +
  ylim(0,30) + 
  theme_minimal() +
  theme(legend.position = "top") 
})

# Arrange the plots in a grid layout (2 rows, 3 columns or adjust as needed)
grid.arrange(grobs = plots, ncol = 3)
```

```{r}
ggplot(cleaned_data, aes(x = etco2, fill = amc)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ hist1) +
  labs(title = "Density of EtCO2 Colored by AMC", subtitle = " Wrapped by History of *insert hist*", fill = "AMC", x = "EtCO2 (mmHg)", y = "Density")

```

### Exploratory/Descriptive Analysis: Main Statistical Analysis:

The p-value (0.5541) for the F-test is larger than the typical
significance threshold of 0.05, meaning the addition of the interaction
terms (amc:hist1 and amc:hist3) does not significantly improve the
model. The RSS difference between the two models is relatively small
(63.984), suggesting that the interactions do not provide a substantial
increase in explanatory power.

However, the Residual sum of squares (RSS) which measures the total
deviation of the predicted values from the actual values, is lower for
Model 2 (5224.8) compared to Model 1 (5288.8), indicating a marginal
improvement in fit when the interaction terms are included.

```{r}
reduced_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + hist4 + hist5, data = cleaned_data)

new_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist3, 
                        data = cleaned_data)

# Compare models, talk about clinical signficiance as why 
anova(reduced_model, new_model)
```

```{r}
reduced_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + hist4 + hist5, data = cleaned_data)

new_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist3, 
                        data = cleaned_data)

# Compare models, talk about clinical signficiance as why 
anova(reduced_model, new_model)
```
### Box Plot Insights
The box plots of EtCO2 by AMC across the different histories illustrate that patients that have compression over the LV have a larger range of EtCO2 values, as the median values around 15-20 mmHg are compare to the median values of those compressed over the LVOT/aortic root at 10-15 mmHg. Across these comorbidities, both the median and IQRs were similar between patients with and without the medical histories. This may suggest that history has a minimal influence on EtCO2 variability.

### Density Plot Observations
The density plots show that there is a slightly higher concentration of low EtCO2 values (from 5-15 mmHg) in the LVOT/aortic root compared to the LV that has a more even distribution from 10-25 mmHg. For reference, awake and healthy adults normally have a range of 35-45 mmHG of EtCO2. There is no clear shift in these density patterns across the different histories, reinforcing both the statistical findings and box plots that medical history does not significantly affect the EtCO2 levels. Though EtCO2 values are similar between those with and without diseases across both LV and LVOT/aortc root, the proportional density differences in the LVOT/aortic root suggests that perfusion impairments in the area impacts EtCO2 dynamics more than LV. Disease that affect the circulation and gas exchange have a more pronounced effect on the aortic root as well. This leads to higher density proportions at lower EtCO2 value. As the statistical tests show no overall significant effect of disease on EtCo2, these patterns may be due to anatomical sensitivity or local perfusion effects than the impact of history. 

#### Density Plot Observations: Understanding the history and EtCO2 levels
Patients with certain pre-existing conditions may already have higher EtCO2 values, independent of compressing the amcs. For those with CAD, CHF, lower EtCO2 values are due to the impaired coronary perfusion, impairing the heart's ability to effectively pump blood and clear CO2 from the body. Those with CKD, the disease has a much higher proportion of the same EtCO2 value compared to those wihtout CKD, reflecting potential renal dysfunction and thus electrolyte imbalances. This causes body fluid and perfusion to not work as effectively, causing a higher density of patients to have lower EtCO2. Those with DM also has a higher density proportion and slighlty higher EtCO2 values at the same EtCO2 values without DM, as the vascular changes alteres CO2 handling and perfusion efficinecy. Interestingly, both people with and without MI have higher densities, though those with a history has slightly higher proportion of low EtCO2 levels, as a previous MI causes lasting damages to the heart's ability to effectively pump blood and clear CO2. 

## Conclusions
### Main Findings: 
Investigating the impact of patient history (including coronary artery disease (CAD), congestive heart failure (CHF), chronic kidney disease (CKD), diabetes mellitus (DM), and myocardial infarcation (MCI)) on EtCO2 levels in both the left ventricle and LVOT/aortic root, our findings revealed that there was no statistically significant difference across all groups. Across the multivariable linear model and ANOVA tests, density plots, and box plots, the results were consistent, suggesting that the patient history does not substantially influence EtCO2 levels. 


### Clinical Significance
Our results suggest that EtCO2 levels are not significantly affected by the different histories, though patients with these different chronic disease may have higher EtCO2 values due to already mentioned compensatory mechanisms. Overall, EtCO2 remains a general marker for cardiac output and blood perfusion during CPR, regardless of the patient's history. As compressions over the LV, compared to LVOT/aortic root, lead to better prognosis with slightly higher EtCO2, optimizing the amc or compression location is crucical for maximizing blood flow and prognosis. 

### Limitations and Future Directions
With approximately 106 valid observations after cleaning the data, the are statistically valid. However, having a larger cohort can provide more representative and generalization insight. In addition, there are some measurement constraints, where EtCO2 was measured during CPR, a highly volatile and unpredictable event. There are many dynamic factors and other confounding factors, such as compression rate, depth, and different ventilation that may introduce some variability that the model does not account for. Future studies in this field can explore how other physiological markers, such as blood pressure or ventilation practices, can examine how real time changes in CPR methodologies influence EtCO2 and prognosis of these patients. 

Overall, though medical history does not statistically influence the relationship between AMC and EtCO2, those with these chronic conditions may show slightly lower or higher EtCO2 values due to its own physiological mechanisms, highlighting the focus on AMC and optimizing the quality of compression during CPR. 


## Citations/References
Baloch K, Rehman Memon A, Ikhlaq U, Umair M, Ansari MI, Abubaker J, Salahuddin N. Assessing the Utility of End-Tidal Carbon Dioxide as a Marker for Fluid Responsiveness in Cardiogenic Shock. Cureus. 2021 Feb 5;13(2):e13164. doi: 10.7759/cureus.13164. PMID: 33692926; PMCID: PMC7938016.

Taghizadieh A, Pouraghaei M, Moharamzadeh P, Ala A, Rahmani F, Basiri Sofiani K. Comparison of end-tidal carbon dioxide and arterial blood bicarbonate levels in patients with metabolic acidosis referred to emergency medicine. J Cardiovasc Thorac Res. 2016;8(3):98-101. doi: 10.15171/jcvtr.2016.21. Epub 2016 Sep 30. PMID: 27777693; PMCID: PMC5075365.

Tepe M, Hakkoymaz H, Kilci Aİ, Gedik MS, Öztürk C, Gökçe MK, Aykan AÇ. Evaluation of the relationship between end-tidal carbon dioxide level and heart failure classification. Rev Assoc Med Bras (1992). 2024 May 20;70(5):e20231499. doi: 10.1590/1806-9282.20231499. PMID: 38775509; PMCID: PMC11110962.

Ölander CH, Vikholm P, Schiller P, Hellgren L. End-Tidal Carbon Dioxide Impacts Brain and Kidney Injury in Experimental Extracorporeal Cardiopulmonary Resuscitation (ECPR). Shock. 2021 Apr 1;55(4):563-569. doi: 10.1097/SHK.0000000000001645. PMID: 32826810.

