---
title: "Final Report"
author: "Group 5: Megan Lui, Michelle Wong, Yael Yossefy, Dirgh Shah"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Introduction

### Scientific Question:

Does the relationship between EtCO2 and AMC behave differently in the
presence of relevant medical history?

### Clinical relevance:

Observing the relationship between EtCO2 (end-tidal carbon dioxide) and
AMC (area of maximum compression) is crucial to understanding the
quality of chest compression during CPR (cardiopulmonary resuscitation),
a life-saving procedure during cardiac arrests. Chest compressions are
crucial to reviving a patient or achieving ROSC or Return to Spontaneous
Circulation, though they requires adequate rate, depth, and location for
the best results. However, CPR guidelines are not as precise, as
differences in body size and histories of chronic disease can alter the
heart's location. When patients have compression in the aortic root,
there is an obstruction of blood flow from the heart and can limit
effective perfusion and worsen the prognosis. Compressions over the left
ventricle or LV can improve the blood circulation and lead to
higher-quality CPR. This relationship between EtCO2 (where higher levels
have been shown to correlate with better blood flow from past studies)
and AMC may differ dependent on a patient's history, as these conditions
can impact the ventilation-perfusion mismatch or other presenting
symptoms.

## Methods

### Design of Study:

The metabolic syndrome and cognitive function study is a retrospective
observational analysis which focuses on investigating the relationship
between EtCO2 and the AMC during CPR and whether this relationship is
affected by the presence of relevant medical histories of patients.

This study focuses on 2 primary areas of compression, the left ventricle
(LV) and the left ventricle outflow tract (LVOT)/aortic root, and how
compression over these AMCs impact EtCO2 levels and whether this impact
changes by the CPR patients' pre existing medical conditions, such as
coronary artery disease, congestive hearth failure, chronic kidney
disease, diabetes mellitus and myocardial infarction.

### Data set Details:

The data set consists of 129 cases of in-hospital cardiac arrests where
transesophageal echocardiography (TEE) imaging was employed during CPR.
This advanced imaging technique is used to identify the specific area of
the chest targeted for compression to precisely classify the AMC - LV or
LVOT.

The dependent variable in the dataset is EtCO2. Measured during CPR,
EtCO2 represents levels of exhaled CO2 and serves as a proxy for cardiac
output. The primary independent variable is AMC, the area of maximum
chest compression and is classified into two categories: left ventricle
(LV) and left ventricular outflow tract (LVOT). Additionally, the
analysis accounts for medical history, captured through binary
indicators for specific conditions such as coronary artery disease
(hist1), congestive heart failure (hist2), chronic kidney disease
(hist3), diabetes mellitus (hist4), and prior myocardial infarction
(hist5).

There are also covariates included in the dataset. These consist of
demographic factors like race and gender, the experience level of the
TEE operator (such as attending physician, resident, etc.), the type of
CPR performed (manual, mechanical, or alternating), and the timing of
cardiac arrest (known or approximated). These variables provide a
comprehensive overview of the factors that might influence EtCO2 during
CPR.

First, the data set underwent pre-processing and cleaning, which
included the removal of invalid data entries. The next step involved
converting the numeric values flagged as outliers (e.g. placeholder
values like 9999) to missing (NA) values and excluding them from the
analysis.

### Cohort:

The study cohort consisted of adult patients who had experienced
in-hospital cardiac arrest and received CPR where TEE was used. The
inclusion criteria required proper TEE imaging data to identify the
precise location of AMC and valid measurements of EtCO2 levels. Patients
with any significant missing data were excluded. After pre-processing
and cleaning, the data set included 106 valid observations (patients)
for further analysis.

```{r}
library(readxl)
library(mosaic)
data <- read_excel("cardiac_arrests.xlsx")
#head(data)
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

data <- data %>%
  mutate(across(where(is.numeric), ~ replace(.x, .x == 9999, NA))) 

data$race <- factor(data$race, levels = c(1, 2, 3, 4, 99), 
                    labels = c("White", "Black/African American", 
                               "Asian", "Latinx", "Other"))
data$amc <- factor(data$amc, levels = c(1, 2), 
                   labels = c("LV", "LVOT/Aortic root"))
data$op_level <- factor(data$op_level, levels = c(1, 2, 3, 99), 
                        labels = c("Attending physician", "Fellow", 
                                   "Resident", "Other"))
data$cpr_type <- factor(data$cpr_type, levels = c(1, 2, 3), 
                        labels = c("Manual", "Mechanical", "Alternating"))
data$hist1 <- factor(data$hist1, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist2 <- factor(data$hist2, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist3 <- factor(data$hist3, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist4 <- factor(data$hist4, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist5 <- factor(data$hist5, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$rosc <- factor(data$rosc, levels = c(0, 1), 
                    labels = c("No", "Yes"))

data$survival <- factor(data$survival, levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$time_approx <- factor(data$time_approx, levels = c(0, 1), 
                           labels = c("Known", "Approximated"))

#summary(data)
cleaned_data <- na.omit(data)
```

### Tables Figures Statistics

First, let's explore the relationship between AMC and EtCO2.

```{r favstats}
knitr::kable(favstats(etco2 ~ amc, data = cleaned_data), col.names = c("Area of Maximum Compression","Min","Q1","Median","Q3","Max","Mean","SD","n","N/A"), caption = "EtCO2 Statistics, grouped by AMC")
```

In our summary table above, we see that the mean EtCO2 levels are higher
for the Left Ventricle AMC (15.75) than for the Aortic Root AMC (12.27).
The same can be said for the median values of EtCO2, which are 15 and
12, respectively.

```{r}
ggplot(cleaned_data, aes(x = etco2, fill = amc)) +
  geom_density(alpha = 0.5) + 
  labs(x = "EtCO2 (mmHg)", y = "Density", fill = "AMC", title = "Density Plot of EtCO2", subtitle = "Colored by Area of Maximum Compression")
```

As we can see by the plot above, the EtCO2 levels are centered about a
higher value, and have a wider spread for the Left Ventricle AMC than
for the Aortic Root AMC, which has a narrower spread and taller peak at
a lower value of EtCO2. This is confirmed by the table above, which
reports a higher EtCO2 Standard Deviation for the Left Ventricle AMC
than the Aortic Root.

Finally, we see all of these findings displayed in our boxplot, of Area
of Maximum Compression along the x-axis, and EtCO2 along the y-axis.

```{r}
ggplot(cleaned_data, aes(x = amc, y = etco2))+
  geom_boxplot(fill = "grey") + 
  labs(x = "Area of Maximal Compression", y = "End-Tidal Carbon Dioxide (mmHg)", title = "BoxPlots of EtCO2 vs. AMC") + 
  ylim(0,30) + 
  theme_minimal()
```

Now let's explore how history of disease affects this relationships:

```{r histories boxplot}
library(gridExtra)

disease_titles <- c("Coronary Artery", "Heart Failure", "Kidney Disease", "Diabetes", "Myocardial Infarction")

plots <- lapply(1:5, function(i) {
  hist_var <- paste0("hist", i)
  ggplot(cleaned_data, aes(x = amc, y = etco2, fill = get(hist_var))) +
    geom_boxplot(outlier.shape = NA) +
    labs(x = "AMC", y = "EtCO2 (mmHg)", title = disease_titles[i], fill = "History") +
    scale_fill_manual(values = c("lightblue", "maroon")) +
    ylim(0, 30) + 
    theme_minimal()+
    theme(legend.position = "bottom")
})

grid.arrange(grobs = plots, ncol = 3)

```

For these plots we removed outliers to visualize the data more clearly.
As we can see, there doesn't seem to be a clear affect on the
relationship by any of the individual histories. We can see this by
comparing the boxplots to the one earlier (before histories were
introduced). We see the same trends, wider boxplots with generally
higher values for the "LV" AMC and thinner, lower boxplots for the
"LVOT/Aortic Root". For Myocardial Infarctions we notice very thin boxes
for those with a history of the disease, and this is because that of our
106 observations in our cleaned data, there were only 4 cases of
Myocardial Infarctions. Thus, there isn't enough data to reach a
conclusion for that plot. In the clean data, there are 20 cases of
Coronary Artery Disease, 14 of Heart Failure, 19 of Kidney Disease, and
40 of Diabetes. There are no clear indications that any of these
diseases affect the relationship between AMC and EtCO2.

To visualize the density plot of EtCO2, colored by AMC, and wrapped by
hist1, Coronary Artery Disease, we see the following:

```{r}
ggplot(cleaned_data, aes(x = etco2, fill = amc)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ hist1) +
  labs(title = "Density of EtCO2 Colored by AMC", subtitle = "Wrapped by History of Coronary Artery Disease", fill = "AMC", x = "EtCO2 (mmHg)", y = "Density")

ggplot(cleaned_data, aes(x = etco2, fill = amc)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ hist3) +
  labs(title = "Density of EtCO2 Colored by AMC", subtitle = "Wrapped by History of Kidney Disease", fill = "AMC", x = "EtCO2 (mmHg)", y = "Density")
```

### Box Plot Insights

The box plots of EtCO2 by AMC across the different histories illustrate
that patients that have compression over the LV have a larger range of
EtCO2 values, as the median values around 15-20 mmHg are compare to the
median values of those compressed over the LVOT/aortic root at 10-15
mmHg. Across these comorbidities, both the median and IQRs were similar
between patients with and without the medical histories. This may
suggest that history has a minimal influence on EtCO2 variability.

### Density Plot Observations

The density plots show that there is a slightly higher concentration of
low EtCO2 values (from 5-15 mmHg) in the LVOT/aortic root compared to
the LV that has a more even distribution from 10-25 mmHg. For reference,
awake and healthy adults normally have a range of 35-45 mmHG of EtCO2.
There is no clear shift in these density patterns across the different
histories, reinforcing both the statistical findings and box plots that
medical history does not significantly affect the EtCO2 levels. Though
EtCO2 values are similar between those with and without diseases across
both LV and LVOT/aortc root, the proportional density differences in the
LVOT/aortic root suggests that perfusion impairments in the area impacts
EtCO2 dynamics more than LV. Disease that affect the circulation and gas
exchange have a more pronounced effect on the aortic root as well. This
leads to higher density proportions at lower EtCO2 value. As the
statistical tests show no overall significant effect of disease on
EtCo2, these patterns may be due to anatomical sensitivity or local
perfusion effects than the impact of history.

#### Density Plot Observations: Understanding the history and EtCO2 levels

Patients with certain pre-existing conditions may already have higher
EtCO2 values, independent of compressing the amcs. For those with CAD,
CHF, lower EtCO2 values are due to the impaired coronary perfusion,
impairing the heart's ability to effectively pump blood and clear CO2
from the body. Those with CKD, the disease has a much higher proportion
of the same EtCO2 value compared to those wihtout CKD, reflecting
potential renal dysfunction and thus electrolyte imbalances. This causes
body fluid and perfusion to not work as effectively, causing a higher
density of patients to have lower EtCO2.

### Analysis Methodology:

A multivariable linear model was implemented to analyse the relationship
between EtCO2 level and AMC. Interaction terms between AMC the the
medical history variables were used in this model to assess whether the
relationship between EtCO2 and AMC varied by the presence of 5 specific
pre-existing medical conditions.

The analysis included two models. First being a reduced model, which
included only the main effects of AMC and medical history. The second is
a full model, which included interaction terms to evaluate whether
medical history affected the relationship between EtCO2 and AMC.

An Analysis of Variance (ANOVA) test was used to compare these models
and evaluate the contribution of the interaction terms to model fit.

### Rationale:

A linear regression model was implemented based on the assumption that
the relationship between EtCO2 and AMC is linear, as CO2 levels in the
blood reflect cardiac output which is influnced by the area of
compression (LV or LVOT). We assume higher EtCO2 levels when the AMC is
LV. This specific approach allows for testing the main effects while
including interaction terms to study more complex relationships. The
inclusion of these interaction terms is very crucial for identifying
whether specific medical histories alter the impact of AMC on EtCO2.

The assumptions include linearity and homoscedasticity, which were
checked via residual plots. Independence was checked using the residual
plots as well. The normality of residuals was assessed through Q-Q
plots. Multicollinearity was evaluated using Variance Inflation Factors
(VIF), with all predictors having suitable values (\< 5).

This methodological approach addresses potential confounding factors and
statistical validity while guaranteeing a thorough investigation of the
research question.

### Assumptions Made:

1.  Linearity and Homoscedasticity (Constant Variance) The relationship
    between the predictors and the outcome are linear. This means that
    changes in the predictors produce proportional changes in the
    outcome.For homoscedasticity, the variance of the residuals should
    be constant across all levels of the independent variables.

```{r}
normal_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist2 + amc:hist3 + 
                          amc:hist4 + amc:hist5, 
                        data = cleaned_data)

plot(normal_model, which = 1)
```

2.  Independence: Looking at the plot above, the residuals from the
    model appear to be independent. The residuals seem to be randomly
    distributed and fluctuate evenly above and below zero. Although
    residual plots alone do not prove independence—they provide evidence
    for it.

3.  Normality of Residuals: The residuals should be normally
    distributed, and we see from the qqplot that they follow the line of
    best fit.

```{r}
plot(normal_model, which = 2)
```

4.  No Perfect Multicollinearity: In regression analysis, it is crucial
    that the independent variables are not highly correlated with one
    another, as this can make it challenging to isolate the individual
    effects of each predictor. While this is typically not an issue when
    the independent variables are categorical or factors in ANOVA, it
    becomes more relevant in multiple regression.

Variance Inflation Factor (VIF) is commonly used to assess
multicollinearity. VIF \< 5 indicates acceptable (low)
multicollinearity. VIF between 5 and 10 signals moderate
multicollinearity and may warrant further investigation. VIF \> 10 is
considered problematic (high multicollinearity), and such values should
generally be avoided.

In our analysis, the VIF values for all variables are well below 5,
which indicates that there is no significant multicollinearity present
in the model. This suggests that the predictor variables are
independent, and we can confidently interpret their individual effects
on the outcome.

```{r}
library(car)
#vif(normal_model, 'predictor')
#vif(normal_model)
library(gt)

results <- data.frame(
  Variable = c("amc", "hist1", "hist2", "hist3", "hist4", "hist5", 
               "amc:hist1", "amc:hist2", "amc:hist3", "amc:hist4", "amc:hist5"),
  Coefficient = c(1.8989, 3.6501, 2.1937, 2.6167, 1.9604, 4.0862, 
                  3.9591, 2.3018, 3.2972, 3.5108, 4.3337)
)

# Create a publication-ready table using gt
results %>%
  gt() %>%
  tab_header(
    title = "Regression Coefficients for Model Variables and Interaction Terms"
  ) %>%
  cols_label(
    Variable = "Variable",
    Coefficient = "Coefficient Estimate"
  ) %>%
  tab_spanner(
    label = "Model Variables and Interactions",
    columns = vars(Variable, Coefficient)
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_column_labels(columns = vars(Variable, Coefficient))
  )

```

### Early linear Regression Model with all possible interaction terms:

Since the assumptions hold, we are able to use a linear regression model
to examine the relationships between our variables. The regression model
we constructed investigates how the area of maximal compression (amc),
history of various medical conditions (hist1 through hist5), and their
interactions affect the end-tidal CO2 (etCO2) levels. We decided to
retain the interaction terms between amc:hist1 (history of coronary
artery disease) and amc:hist3 (history of chronic kidney disease) based
on clinical reasoning, despite their lack of statistical significance.
These interactions were included to explore potential nuanced effects on
etCO2 levels that could be clinically meaningful. Although the
coefficients for these interactions were not statistically significant,
their retention helps ensure a comprehensive understanding of the
relationships at play.

```{r}
library(gt)
library(dplyr)
library(broom)
library(kableExtra)
#summary(normal_model)

model_summary <- tidy(normal_model)
normal_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + hist4 + hist5 + 
                   amc:hist1 + amc:hist2 + amc:hist3 + amc:hist4 + amc:hist5, 
                   data = cleaned_data)

# Tidy the model output using broom
model_summary <- tidy(normal_model)

# Add significance stars manually
model_summary <- model_summary %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ " "
    )
  )

# Create the table with kableExtra
model_summary %>%
  select(term, estimate, std.error, statistic, p.value, significance) %>%
  kable("html", caption = "Regression Results for ETCO2") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                font_size = 12) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "2cm") %>%
  column_spec(5, width = "2cm") %>%
  column_spec(6, width = "1.5cm")
```

### Early linear Regression Model with all possible interaction terms:

Since the assumptions of ANOVA hold (Independence, multicollinearity,
normality of residuals, linearity, and homoscedasticity), we are able to
use a linear regression model to examine the relationships between our
variables.

```{r}
#summary(normal_model)
```

## Results

### Exploratory/Descriptive Analysis: Main Statistical Analysis:

The p-value (0.5541) for the F-test is larger than the typical
significance threshold of 0.05, meaning the addition of the interaction
terms (amc:hist1 and amc:hist3) does not significantly improve the
model. The RSS difference between the two models is relatively small
(63.984), suggesting that the interactions do not provide a substantial
increase in explanatory power.

However, the Residual sum of squares (RSS) which measures the total
deviation of the predicted values from the actual values, is lower for
Model 2 (5224.8) compared to Model 1 (5288.8), indicating a marginal
improvement in fit when the interaction terms are included.

```{r}
reduced_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + hist4 + hist5, data = cleaned_data)

new_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist3, 
                        data = cleaned_data)

# Compare models, talk about clinical signficiance as why 
anova(reduced_model, new_model)
```

```{r}
reduced_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + hist4 + hist5, data = cleaned_data)

new_model <- lm(etco2 ~ amc + hist1 + hist2 + hist3 + 
                          hist4 + hist5 + amc:hist1 + amc:hist3, 
                        data = cleaned_data)

# Compare models, talk about clinical signficiance as why 

```

## Conclusions

### Main Findings:

Investigating the impact of patient history (including coronary artery
disease (CAD), congestive heart failure (CHF), chronic kidney disease
(CKD), diabetes mellitus (DM), and myocardial infarcation (MCI)) on
EtCO2 levels in both the left ventricle and LVOT/aortic root, our
findings revealed that there was no statistically significant difference
across all groups. Across the multivariable linear model and ANOVA
tests, density plots, and box plots, the results were consistent,
suggesting that the patient history does not substantially influence
EtCO2 levels.

### Clinical Significance

Our results suggest that EtCO2 levels are not significantly affected by
the different histories, though patients with these different chronic
disease may have higher EtCO2 values due to already mentioned
compensatory mechanisms. Overall, EtCO2 remains a general marker for
cardiac output and blood perfusion during CPR, regardless of the
patient's history. As compressions over the LV, compared to LVOT/aortic
root, lead to better prognosis with slightly higher EtCO2, optimizing
the amc or compression location is crucial for maximizing blood flow and
prognosis.

### Limitations and Future Directions

With approximately 106 valid observations after cleaning the data, the
are statistically valid. However, having a larger cohort can provide
more representative and generalization insight. In addition, there are
some measurement constraints, where EtCO2 was measured during CPR, a
highly volatile and unpredictable event. There are many dynamic factors
and other confounding factors, such as compression rate, depth, and
different ventilation that may introduce some variability that the model
does not account for. Future studies in this field can explore how other
physiological markers, such as blood pressure or ventilation practices,
can examine how real time changes in CPR methodologies influence EtCO2
and prognosis of these patients.

Overall, though medical history does not statistically influence the
relationship between AMC and EtCO2, those with these chronic conditions
may show slightly lower or higher EtCO2 values due to its own
physiological mechanisms, highlighting the focus on AMC and optimizing
the quality of compression during CPR.

## Citations/References

Baloch K, Rehman Memon A, Ikhlaq U, Umair M, Ansari MI, Abubaker J,
Salahuddin N. Assessing the Utility of End-Tidal Carbon Dioxide as a
Marker for Fluid Responsiveness in Cardiogenic Shock. Cureus. 2021 Feb
5;13(2):e13164. doi: 10.7759/cureus.13164. PMID: 33692926; PMCID:
PMC7938016.

Taghizadieh A, Pouraghaei M, Moharamzadeh P, Ala A, Rahmani F, Basiri
Sofiani K. Comparison of end-tidal carbon dioxide and arterial blood
bicarbonate levels in patients with metabolic acidosis referred to
emergency medicine. J Cardiovasc Thorac Res. 2016;8(3):98-101. doi:
10.15171/jcvtr.2016.21. Epub 2016 Sep 30. PMID: 27777693; PMCID:
PMC5075365.

Tepe M, Hakkoymaz H, Kilci Aİ, Gedik MS, Öztürk C, Gökçe MK, Aykan AÇ.
Evaluation of the relationship between end-tidal carbon dioxide level
and heart failure classification. Rev Assoc Med Bras (1992). 2024 May
20;70(5):e20231499. doi: 10.1590/1806-9282.20231499. PMID: 38775509;
PMCID: PMC11110962.

Ölander CH, Vikholm P, Schiller P, Hellgren L. End-Tidal Carbon Dioxide
Impacts Brain and Kidney Injury in Experimental Extracorporeal
Cardiopulmonary Resuscitation (ECPR). Shock. 2021 Apr 1;55(4):563-569.
doi: 10.1097/SHK.0000000000001645. PMID: 32826810.
