---
title: "biostats"
author: "Group 5"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
library(readxl)
cardiac_data <- read_excel("cardiac_arrests.xlsx")
head(data)
```

### Cleaning

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

cardiac_data <- cardiac_data %>%
  mutate(across(everything(), ~ replace(.x, .x == 9999, NA))) %>%
  na.omit()
```

### EDA

```{r}
# AMC Distribution by Medical History
cardiac_data %>%
  select(amc, starts_with("hist")) %>%
  pivot_longer(cols = starts_with("hist"), names_to = "Medical_History", values_to = "Presence") %>%
  filter(!is.na(Presence)) %>%
  group_by(amc, Medical_History) %>%
  summarise(Count = sum(Presence, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(amc, labels = c("LV", "LVOT/Aortic Root")), y = Count, fill = Medical_History)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Medical History Counts by AMC Type", x = "AMC Type", y = "Count") +
  theme_minimal()

# Boxplot of EtCO2 by AMC and Medical History
ggplot(cardiac_data, aes(x = factor(amc, labels = c("LV", "LVOT/Aortic Root")), y = etco2, fill = factor(hist1))) +
  geom_boxplot() +
  labs(title = "EtCO2 by AMC and History of Coronary Artery Disease (hist1)",
       x = "AMC Type",
       y = "EtCO2",
       fill = "History of CAD (hist1)") +
  theme_minimal()
```

### ANOVA

```{r}
# Linear model with interactions
interaction_model <- lm(etco2 ~ amc * (hist1 + hist2 + hist3 + hist4 + hist5), data = cardiac_data)

# Summary of the model
summary(interaction_model)

# ANOVA to test interaction significance
anova(interaction_model)

# Check the residuals for model assumptions
par(mfrow = c(2, 2))  # Display diagnostic plots
plot(interaction_model)
```

## Introduction

### Scientific Question:

Does the relationship between EtCO2 and AMC behave differently in the
presenceof relevant medical history?

### Clinical relevance:

Observing the relationship between EtCO2 (end-tidal carbon dioxide) and
AMC (area of maximum compression) is crucial to understanding the
quality of chest compression during CPR (cardiopulmonary resuscitation),
a life-saving procedure during cardiac arrests. Chest compressions are
crucial to reviving a patient or achieving ROSC or Return to Spontaneous
Circulation, though they requires adequate rate, depth, and location for
the best results. However, CPR guidelines are not as precise, as
differences in body size and histories of chronic disease can alter the
heart's location. When patients have compression in the aortic root,
there is an obstruction of blood flow from the heart and can limit
effective perfusion and worsen the prognosis. Compressions over the left
ventricle or LV can improve the blood circulation and lead to
higher-quality CPR. This relationship between EtCO2 (where higher levels
have been shown to correlate with better blood flow from past studies)
and AMC may differ dependent on a patient's history, as these conditions
can impact the ventilation-perfusion mismatch or other presenting
symptoms.

## Methods

Design of Study: 
Data Set Details: 
Cohort: 

Chosen Analysis Methodology: lgm 
Rationale: We cannot assume a linear relationship with medical history 

## Results

Tables Figures Statistics

Exploratory/Descriptive Analysis: Main Statistical Analysis:

## Conclusions

Main Findings: Answer Scientific Question: Limitations:

##References

Citations References
