---
title: "biostats"
author: "Group 5"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
library(readxl)
cardiac_data <- read_excel("cardiac_arrests.xlsx")
head(data)
```

### Cleaning 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

data[data == 9999] <- NA

data$race <- factor(data$race, levels = c(1, 2, 3, 4, 99), 
                    labels = c("White", "Black/African American", 
                               "Asian", "Latinx", "Other"))

data$amc <- factor(data$amc, levels = c(1, 2), 
                   labels = c("LV", "LVOT/Aortic root"))

data$op_level <- factor(data$op_level, levels = c(1, 2, 3, 99), 
                        labels = c("Attending physician", "Fellow", 
                                   "Resident", "Other"))

data$cpr_type <- factor(data$cpr_type, levels = c(1, 2, 3), 
                        labels = c("Manual", "Mechanical", "Alternating"))

data$hist1 <- factor(data$hist1, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist2 <- factor(data$hist2, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist3 <- factor(data$hist3, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist4 <- factor(data$hist4, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$hist5 <- factor(data$hist5, levels = c(0, 1), 
                     labels = c("No", "Yes"))

data$rosc <- factor(data$rosc, levels = c(0, 1), 
                    labels = c("No", "Yes"))

data$survival <- factor(data$survival, levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$time_approx <- factor(data$time_approx, levels = c(0, 1), 
                           labels = c("Known", "Approximated"))

summary(data)
sum(is.na(data))
```

### Check cleanliness
```{r}
cleaned_data <- na.omit(data)

sum(is.na(cleaned_data))
```

### EDA 
```{r}
# AMC Distribution by Medical History
cardiac_data %>%
  select(amc, starts_with("hist")) %>%
  pivot_longer(cols = starts_with("hist"), names_to = "Medical_History", values_to = "Presence") %>%
  filter(!is.na(Presence)) %>%
  group_by(amc, Medical_History) %>%
  summarise(Count = sum(Presence, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(amc, labels = c("LV", "LVOT/Aortic Root")), y = Count, fill = Medical_History)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Medical History Counts by AMC Type", x = "AMC Type", y = "Count") +
  theme_minimal()

# Boxplot of EtCO2 by AMC and Medical History
ggplot(cardiac_data, aes(x = factor(amc, labels = c("LV", "LVOT/Aortic Root")), y = etco2, fill = factor(hist1))) +
  geom_boxplot() +
  labs(title = "EtCO2 by AMC and History of Coronary Artery Disease (hist1)",
       x = "AMC Type",
       y = "EtCO2",
       fill = "History of CAD (hist1)") +
  theme_minimal()
```

### More plots
```{r}
library(ggplot2)

# hist1
ggplot(cleaned_data, aes(x = etco2, color = hist1)) +
  geom_density() +
  facet_wrap(~ amc) +
  labs(title = "Relationship between EtCO2 and AMC by Coronary Artery Disease History",
       x = "EtCO2 (mmHg)", y = "Density") +
  theme_minimal()
```

```{r}
# hist2
ggplot(cleaned_data, aes(x = etco2, color = hist2)) +
  geom_density() +
  facet_wrap(~ amc) +
  labs(title = "Relationship between EtCO2 and AMC by Congestive Heart Failure History",
       x = "EtCO2 (mmHg)", y = "Density") +
  theme_minimal()
```

```{r}
# hist3
ggplot(cleaned_data, aes(x = etco2, color = hist3)) +
  geom_density() +
  facet_wrap(~ amc) +
  labs(title = "Relationship between EtCO2 and AMC by ",
       x = "EtCO2 (mmHg)", y = "Density") +
  theme_minimal()
```

```{r}
# hist4
ggplot(cleaned_data, aes(x = etco2, color = hist4)) +
  geom_density() +
  facet_wrap(~ amc) +
  labs(title = "Relationship between EtCO2 and AMC by ",
       x = "EtCO2 (mmHg)", y = "Density") +
  theme_minimal()
```

```{r}
# hist5
ggplot(cleaned_data, aes(x = etco2, color = hist2)) +
  geom_density() +
  facet_wrap(~ amc) +
  labs(title = "Relationship between EtCO2 and AMC",
       x = "EtCO2 (mmHg)", y = "Density") +
  theme_minimal()
```


### ANOVA
```{r}
interaction_model <- glm(amc ~ etco2 * hist1 + etco2 * hist2 + etco2 * hist3 + 
                         etco2 * hist4 + etco2 * hist5, 
                         data = cleaned_data, family = binomial)

summary(interaction_model)
```


## Introduction

### Scientific Question: 
Does the relationship between EtCO2 and AMC behave differently in the presenceof relevant medical history? 

### Clinical relevance:
Observing the relationship between EtCO2 (end-tidal carbon dioxide) and
AMC (area of maximum compression) is crucial to understanding the
quality of chest compression during CPR (cardiopulmonary resuscitation),
a life-saving procedure during cardiac arrests. Chest compressions are
crucial to reviving a patient or achieving ROSC or Return to
Spontaneous Circulation, though they requires adequate rate, depth, and
location for the best results. However, CPR guidelines are not as
precise, as differences in body size and histories of chronic disease
can alter the heart's location. When patients have compression in the
aortic root, there is an obstruction of blood flow from the heart and
can limit effective perfusion and worsen the prognosis. Compressions
over the left ventricle or LV can improve the blood circulation and lead
to higher-quality CPR. This relationship between EtCO2 (where higher
levels have been shown to correlate with better blood flow from past
studies) and AMC may differ dependent on a patient's history, as these
conditions can impact the ventilation-perfusion mismatch or other
presenting symptoms.

## Methods

Design of Study:
Data Set Details:
Cohort:
Chosen Analysis Methodology:
Rationale:

## Results

Tables
Figures
Statistics

Exploratory/Descriptive Analysis:
Main Statistical Analysis:

## Conclusions

Main Findings:
Answer Scientific Question:
Limitations:

##References

Citations
References