---
title: "biostats"
author: "Group 5"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
library(readxl)
cardiac_data <- read_excel("cardiac_arrests.xlsx")
head(data)
```

### Cleaning 
```{r}
library(dplyr)

cardiac_data <- cardiac_data %>%
  mutate(across(everything(), ~ replace(.x, .x == 9999, NA)))
```

### Clustering 
```{r}

```

## Introduction

Scientific Question: Does the relationship between EtCO2 and AMC behave differently in the presence
of relevant medical history? 
Clinical relevance:

## Methods

Design of Study:
Data Set Details:
Cohort:
Chosen Analysis Methodology:
Rationale:

## Results

Tables
Figures
Statistics

Exploratory/Descriptive Analysis:
Main Statistical Analysis:

## Conclusions

Main Findings:
Answer Scientific Question:
Limitations:

##References

Citations
References