---
title: "biostats"
author: "Group 5"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 2
    number_sections: no
    theme: cerulean
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
---

```{r,echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r}
library(readxl)
cardiac_data <- read_excel("cardiac_arrests.xlsx")
head(data)
```

### Cleaning 
```{r}
library(dplyr)

cardiac_data <- cardiac_data %>%
  mutate(across(everything(), ~ replace(.x, .x == 9999, NA)))

```

### Clustering 
```{r}

```


### Fit logistic regression model 
```{r}
# Load necessary packages
library(readxl)
library(dplyr)

# 2. Logistic Regression: Predicting ROSC
glm_rosc <- glm(rosc ~ age + sex + race + amc + etco2 + cpr_type + time_to_determination +
                  hist1 + hist2 + hist3 + hist4 + hist5,
                data = cardiac_data, family = binomial)

summary(glm_rosc)

# 3. Logistic Regression: Predicting Survival
glm_survival <- glm(survival ~ age + sex + race + amc + etco2 + cpr_type + time_to_determination +
                      hist1 + hist2 + hist3 + hist4 + hist5,
                    data = cardiac_data, family = binomial)

summary(glm_survival)
```

### Coefficients & Odds Ratios 
```{r}
exp(coef(logit_model))
```

### Model evaluation 
```{r}
library(pROC)

# Predicted probabilities
pred_probs <- predict(logit_model, type = "response")

# AUC
roc_curve <- roc(data_clean$survival, pred_probs)
auc(roc_curve)
```

## Introduction

Scientific Question: Does the relationship between EtCO2 and AMC behave differently in the presence
of relevant medical history? 
Clinical relevance:

## Methods

Design of Study:
Data Set Details:
Cohort:
Chosen Analysis Methodology:
Rationale:

## Results

Tables
Figures
Statistics

Exploratory/Descriptive Analysis:
Main Statistical Analysis:

## Conclusions

Main Findings:
Answer Scientific Question:
Limitations:

##References

Citations
References